The variability of Betelgeuse explained by surface convection\thanks{Based on observations obtained at the T\'elescope Bernard Lyot (TBL) at Observatoire du Pic du Midi, CNRS/INSU and Universit\'e de Toulouse, France.}

### 2024-09-04 main.tex
Add files via upload

    \author{{ Q.~Pilate}\inst{1},{ A.~L{\'o}pez Ariste}\inst{1},{ A.~Lavail}\inst{1},{ Ph. Mathias}\inst{1} }

@@

\section{Introduction}
\label{sec: Introduction}
Betelgeuse is a prototypical red supergiant (RSG), known to be a semi-regular variable. Several periods can be found in the 
literature, usually clustered around the so-called long secondary period (LSP) of 2000 d, plus shorter periods around 400 and 200 d \citep{kiss_variability_2006}. The origin of the LSP remains a source of discussion: some have tried to find the explanation of the LSP in the lifetime of the large convective cells present at the surface of RSGs \citep[e.g][]{stothers_giant_2010} while others have invoked a magnetic field \citep{wood_long_2004}. The shorter periods of 400 and 200 d are given a different origin than the LSP. By studying a large sample of RSGs, \cite{kiss_variability_2006} attributed the 400 d period to radial pulsation modes, while the 200 d period being attributed to the first overtone \citep{joyce_standing_2020}, in agreement with the historical work of \cite{stothers_pulsation_1969}.\\

Besides radial pulsation, surface activity leading to random brightness variation has also been mentioned to explain the variability of Betelgeuse, first theorised by \cite{schwarzschild_scale_1975}. More recently, \cite{gray_mass_2008} proposed that the 400 d period was a consequence of convective activity. Betelgeuse is known to present large convective cells with lifetimes of the order of one to two years \citep{lopez_ariste_convective_2018} and measurable changes 
within the span of one week. Bright convection cells near the disk center will increase the integrated brightness of the star compared with other situations where such cells are found near the edges. So, even without invoking the formation of 
dust or other changes in opacity, one may argue that the simple evolution of convective patterns of the star may create variability. Such 
variability could be expected to be random, but will present quasi-periodicities related to the typical time scales of the 
convective patterns \citep{gray_mass_2008}. \\

%more or less well defined 
%periods around 400 and 200 days. 
%Such periods are often linked to radial pulsation modes,   and have been used to model 
%the characteristics of the resonant cavity, hence the size of the star and eventually its evolutionary stage \citep{saio_evolutionary_2023}.

At the end of 2019, Betelgeuse reached a historical minimum in its luminosity, called the Great Dimming \citep{guinan_fall_2020}. 
From interferometric and spectroscopic data, it has been proposed that this event was caused by the formation of a cloud of dust close to the line of sight \citep{montarges_dusty_2021}. 
Interferometric images show a drop in luminosity in the southern hemisphere of Betelgeuse, which could be caused by a mass loss event and leading to this dimming \citep{dupree_great_2022}. Other hypotheses have also been explored, such as a drop in temperature \citep{harper_photospheric_2020}, or an increase in the molecular opacity \citep{kravchenko_atmosphere_2021}. In this event, it turns out that a change in brightness was not due to pulsation but to a change in the 
brightness distribution over the stellar disk.
%due to the presence of large quantities of dust.
Since the end of the Great Dimming event, Betelgeuse has continued its random variation of brightness. 
But quite interestingly, 
it has been shown by \cite{jadlovsky_analysis_2023} and \cite{dupree_great_2022} that the periodicity of Betelgeuse has changed since the dimming. Using the light curves from AAVSO, 
the cited authors showed that before the dimming, the dominant period of Betelgeuse was the 400 d period, while after the great dimming, the main periods of Betelgeuse have shortened, oscillating between 97 d and 230 d \citep{dupree_great_2022}, revealing a change in the behaviour of the atmosphere of Betelgeuse.
While the great dimming may be seen as 
a singular event, such modifications of the variability periods bring up the question of whether all changes in brightness can be due to similar changes in the brightness 
distribution over the disk and unrelated to any pulsation phenomenon. 
Our interest in an alternative explanation of the variability
in terms  of convective patterns arose. In order to address this question, we seek such typical periods of 400 and 200 d in  
observational proxies related to the convective
activity but not to any pulsation, such as the linear polarization
spectra.\\
%but unrelated to pulsations and which cannot be directly linked to brightness.
%Faced with the dramatic change in stellar parameters and age that such change in a pulsation period would 
%require,

@@

Linear polarization observed in the atomic lines is therefore a proxy of convection, a priori unrelated to radial 
pulsations but linked to the brightness inhomogeneities due to the convective patterns in Betelgeuse. If one can find the aforementioned variability periods 
in linear polarization, we can conclude that these periods are related to the convective activity which originates the linear polarization signals.
%and unrelated to any radial pulsation phenomena. 
This is the purpose of the present work.\

In section \ref{Section 2}, we describe the dataset of linear polarization obtained with Narval and Neo-Narval at the 2-meter Télescope Bernard Lyot (TBL), as well as the Lomb-Scargle periodograms used to derive periods. In section \ref{section 3}, we seek periods in the LSD profiles of Betelgeuse using the Lomb-Scarlge periodogram. In section \ref{section 4}, we associate the 200 d period with the convective timescale of the smallest granules and the 330 d periodicity with the largest granules. We speculate on an explanation in the change of variability of Betelgeuse before and after the great dimming.

@@

Betelgeuse has been observed for the last 10 years with Narval and Neo-Narval at the Telescope Bernard Lyot\footnote[1]{https://tbl.omp.eu/}. These two instruments measure the polarization 
over the visible and near infrared spectra of Betelgeuese (390-1000 nm) with high spectral resolution (R=65000) and high polarimetric sensitivity. The signal-to-noise ratios are not sufficient to measure the weak polarization signals in individual atomic lines of the 
spectrum. These amplitudes are known to be of the order of $10^{-4}$ times the continuum intensity, and they only exceptionally reach amplitudes of $10^{-3}$. When these high amplitudes are present, enough photons can be accumulated per spectral bin to 
detect the linear polarization signal above noise in lines \citep{auriere_discovery_2016}. Most commonly, the amplitudes of linear polarization are below noise levels. On those 
occasions, we have to add up the signals of thousands of lines to increase the signal-to-noise ratios. This addition 
is performed through the Least-Squares Deconvolution (LSD) \citep[LSD;][]{donati_spectropolarimetric_1997,kochukhov_least-squares_2010}.% and assumes that the 
%linear polarization signal is similar 
%in all spectral lines up to scale factors in the sampling and amplitude.
This technique has been successfully used in the past to measure 
magnetic field distributions over stellar surfaces \citep[e.g][]{donatiLandstreet2009,reiners2012,kochukhov2021} and is now employed to produce images of the brightness distributions in the photosphere 
of Betelgeuse and other RSGs, as cited in section~\ref{sec: Introduction}.

@@

%\subsection{Lomb-Scargle periodograms of the profiles}

@@

Attempting to identify periods in an astrophysical context faces the challenge of unevenly spaced observed data. To overcome this hurdle, 
we used the Lomb-Scargle periodogram \citep{lomb_least-squares_1976,scargle_studies_1982}. This method involves fitting sinusoids to the data using least-squares with frequency sampled between the first and last observation. The quality of the fit determines the power attributed to a frequency. We applied this technique to the  LSD profiles of Stokes $I$, $U$, $Q$ and to the total linear polarization of Betelgeuse observed by the TBL over a span of 10 years (2014-2024). However, the dataset available post-dimming is insufficient to produce meaningful periodograms. Therefore, our focus was on periodograns computed exclusively from data obtained before the great dimming, covering the period from 2014 to the end of 2019.

@@

We first computed the Lomb-Scargle periodogram at each wavelength of the LSD profile, ranging from -25~\kms to +60~\kms with a step of 1.8~\kms to cover the entire 
signal present in the LSD profile. \cite{lopez_ariste_convective_2018} identified the most blueshifted signal towards -20~\kms and interpreted it as  the maximum velocity of the rising plasma. The most redshifted signal was often found at +40~\kms, interpreted 
as the rest velocity of the star. Sometimes, signals at velocities greater than +40~\kms can be found, which can be interpreted either as dark and cold plasma falling back to Betelgeuse or as rising plumes of plasma located on the hidden face of Betelgeuse, ascending so high that they appear above the limb \citep[for the case of the RSG $\mu$~Cep]{lopez_ariste_height_2023}.

% The LS periodogram  computed for each frequency 
% in the LSD profile must show periodicities around the LSP, but also on the 400 days and 200 days period. If the LS periodogram finds a periodicity at a
% given wavelength, it would appear as a strong LS power. However, we expect a change in the periodicity after the dimming. 
% This would lead to a decrease in the Lomb-Scargle power. By comparing LS periodogram before and after the dimming, we expect to see a change in the LS power,
% especially at the 400 days period.

Figures \ref{LS intensity},\ref{LS Q},\ref{LS U} and \ref{LS linear polarization} show the Lomb-Scargle periodogram of the LSD profile for Stokes $I$, Stokes $Q$, Stokes $U$ and the total linear polarization, respectively. 
%The abscissa is the frequency (in $\mathrm{days^{-1}}$), while the ordinate is the heliocentric radial velocity in km/s. 
In each figure, the upper panel depicts the Lomb-Scargle periodogram of each velocity bin (black line) along with the total average (red line). The green line represents the window function. The lower panel shows the Lomb-Scargle periodogram for each velocity bin, with the white dotted lines marking the 400 and 200 d periods for reference. The right panel represents the mean profile of each Stokes parameter at each velocity bin.

\begin{figure}[!h]
    \centering
    \includegraphics[width=0.5\textwidth]{Lomb-Scargle Intensity.pdf}
    \caption{Lomb-Scargle periodogram of the LSD profile of intensity.
    The upper panel is the Lomb-Scargle periodogram for each velocity bin (grey lines) and the average (red line). The green line is the window function.
    In the lower panel, the two white dashed lines mark respectively the 400 d and the 200 d periods. The right panel is the average intensity profile for each velocity bin.}
    \label{LS intensity}
\end{figure}

@@

%\section{Recovering the variability of Betelgeuse in the LSD profile}

%\subsection{Intensity}
Examining Fig.\ref{LS intensity}, which shows the Lomb-Scargle periodogram of the LSD of the intensity profile, we observe that the 400 and 200 d periods seem to be captured by the periodogram. However, the 400 d period is uncomfortably close to the peak of the window function. Furthermore, both periods are located at the same HRV and in the blue wing of the profile. Interestingly, \cite{mathias_evolution_2018} previously found this 200 d periodicity in spectroscopic observation, despite of a shorter observation period.

Figures \ref{LS Q} and \ref{LS U} display the Lomb-Scargle periodograms of Stokes $Q$ and $U$. These periodograms exhibit significant differences compared to the intensity one. In Fig. \ref{LS Q}, a prominent signal is evident around 200 d. Regarding Stokes $U$ in Fig. \ref{LS U}, it appears that the primary period is approximately $0.003 \ \mathrm{d^{-1}}$ (equivalent to 330 d). Other periods, such as those at 200 d or 250 d ($0.004 \ \mathrm{d^{-1}}$) are present, but are difficult to trust. From both periodograms of Stokes $Q$ and $U$, we recover the 200 d period, and also the 330 d period is notable, which aligns closely with the 400 d period reported in the literature and is in agreement with the timescale of the hysteresis loop reported by \cite{kravchenko_tomography_2019}.  
Figure \ref{LS linear polarization} shows the Lomb-Scargle periodogram of $\sqrt{Q^2+U^2}$, representing the total linear polarization of Betelgeuse. 
This periodogram confirms the significant powers at both 330 d and 200 d, consistent with the periodograms of Stokes $Q$ and $U$.

%Nonetheless, the power of this period peaks at a HRV of around 40 km/s. We recall that this velocity is the rest velocity of the star. 
%Every signal beyond this velocity is either due to  plasma sinking in Betelgeuse, or to plumes of plasma rising beyond and above the limb of the star. 
%Such plumes from the back hemisphere have been often seen in other red supergiants like $\mu$ Cep \citep{lopez_ariste_height_2023}. 
%This suggests that both signals at these period of 400 days found in intensity and total linear polarisation 
%could well be due to the periodic appearance of plumes in the back hemisphere rising high enough to appear above
%the visible limb and generating a supplementary source of brightness. Once again we find a plausible explanation 
%of these periods unrelated to pulsations, but rather to the convective dynamics of the star.

% Up to this point, we want to pinpoint the two elements that are a clue 
% to understand the variability of Betelgeuse: the 400 days period is present in the linear polarization signal and also at a HRV above 40 km/s.
% In any case, from the linear polarization and the intensity profile, we recover the different periods of Betelgeuse. Since linear polarization is 
% associated to convection and because we recover the periods of Betelgeuse, the main mechanism behind the variability of Betelgeuse points toward surface convection.

@@

Betelgeuse is observed, on average, every month by the TBL, enabling the tracking of its surface activity through this technique of polarimetric imaging. This technique has previously allowed for the estimation of  the size and the
 lifetime of convective cells on the surface \citep{lopez_ariste_convective_2018}. From these images, we computed a photo-center, a quantity sensitive to the size and number of convective cells. 
A homogeneous star will have a photo-center displacement coinciding with the barycenter of the star, whereas a star with one or two large convective cells will exhibit a more siginificant photo-center displacement, up to a few percent of the stellar radius in the case of RSGs \citep{chiavassa_probing_2022}.

Since the photo-center is linked to surface convection, it is worth checking for periods in its 
dynamics over the 5 years of observations of Betelgeuse before the dimming. While these periods may overlap with those presented in the previous section, they are likely to capture additional aspects of the phenomena at work.

%We computed the Lomb-Scargle periodogram of the displacement of the photo-center.
However, before proceeding to search for periods in the displacement of the photo-center, it is important to address a key issue regarding the interpretation of such images. 
Linear polarization suffers from a $180 ^\circ$ degree ambiguity, as mentioned in \cite{auriere_discovery_2016}. Consequently, our images 
can be 
rotated by $180^\circ$ degree, and the brightness distribution will still fit the observed LSD polarization profiles. 
If each observation were treated independently, the algorithm's solution could be any of the possible ambiguous solutions. To ensure continuity 
between the image series, 
for a given day, we use the brightness distribution of the previous day as the initial point 
of the  fitting iteration. The first image in the series begins its minimisation iteration with a random brightness distribution. 
Although we can produce a series of consistent images, it is important to keep in mind that the photo-center displacement computed from the series 
will be affected by the initial image. 
To overcome this issue, we computed
the photo-center displacement from 100 different time series, each starting from a different initial image. This approach aims to recover ensemble properties independent of the choice of the first image.

\begin{figure}[!h]
    \centering
    \includegraphics[width=0.5\textwidth]{Lomb-Scargle Photo-center.pdf}
    \caption{Lomb-Scargle periodogram of the 100 photo-center displacement series of Betelgeuse. Each blue lines correspond to a Lomb-Scargle periodogram of one photo-center
     displacement. The black line is the average of the 100 periodogram. The red dashed lines represent the 400 and 200 d period respectively.
     The green line is the window function. The black horizontal line marks the 2$\sigma$ confidence level.}
    \label{LS photocenter}
\end{figure}

Figure \ref{LS photocenter} shows the Lomb-Scargle periodogram of the photo-center displacement for each of the 100 series (blue lines) and the average Lomb-Scargle periodogram (black line). 
%As before, the upper panel shows the Lomb-Scargle periodogram from data before the great dimming only, whereas we used the whole set of data for the lower panel.
The red dashed lines indicate the 400 and 
200 d periods, respectively. The black horizontal line represents the 2$\sigma$ confidence level. Similar to previous figures, the window function is represented by the green line. Interestingly, we find two peaks around the 400 d period in the periodogram, one at 0.0024 $\mathrm{d^{-1}}$ and one at 0.003 $\mathrm{d^{-1}}$ which are above the 2$\sigma$ confidence level. Although they are not individually significant to confirm the presence of such period, these peaks are consistent with the findings of the previous section. Since polarimetry imaging involves only surface convection, this provides further support for a variability being explained by surface convection alone.

@@

In Fig.\ref{Lomb Scargle AAVSO}, we computed the periodogram from the light curve of Betelgeuse from 1990 to 2024 (blue line), alongside the periodogram computed exclusively with AAVSO observations made less than 5 days from our observation dates of the TBL (orange line). 
For the periodogram since 1990, we binned the observations with an interval of 10 days, as Betelgeuse has been more observed in the 21st century \citep{kiss_variability_2006}. This periodogram  recovers the 400 d period and also a period close to 200 d. Other visible peaks are rather 
due to the windowing effect. However, when focusing on the periodogram derived from the light curve corresponding to the TBL observation dates, we fail to retrieve the periods mentioned in the literature. The periodicities found in the TBL observations gain validity through this comparison: they were evident in the TBL data, but barely recovered in the AAVSO data when pruned to the same observing dates. We can anticipate that extended spectropolarimetric observations 
of Betelgeuse will make these periodicities seen in linear polarization. Due to convective structures, these periodicities will be clearer and clearer.

\begin{figure}[!h]
    \centering
    \includegraphics[width=0.5\textwidth]{Lomb-Scargle AAVSO.pdf}
    \caption{Lomb-Scargle periodogram of the light curve of Betelgeuse since 1990 as retrived from the AAVSO database (blue line). When only dates corresponding to TBL observations are retained in that light curve, the periodogram in orange color is obtained. The two black dashed lines are respectively the 400 and 200 d periods for visual reference.}
    \label{Lomb Scargle AAVSO}
\end{figure}

>>>>

    \author{{ Q.~Pilate}\inst{1},{ A.~López Ariste}\inst{1},{ A.~Lavail}\inst{1},{ Ph. Mathias}\inst{1} }

@@

\section{Introduction}
\label{sec: Introduction}
Betelgeuse is a prototypical red supergiant (RSG), known to be a semi-regular variable. Several periods can be found in the 
literature, usually clustered around the so-called long secondary period (LSP) of 2000 d, plus shorter periods around 400 and 200 d \citep{kiss_variability_2006}. The origin of the LSP remains a source of discussion: some have tried to find the explanation of the LSP in the lifetime of the large convective cells present at the surface of RSGs \citep[e.g][]{stothers_giant_2010} while others have invoked a magnetic field \citep{wood_long_2004}. \textbf{The shorter periods of 400 and 200 d are given a different origin than the LSP, and are traditionally attributed to radial pulsation driven by the $\kappa$-mechanism, the 400 d period  being the fundamental mode and the 200 d the first overtone. By studying a large sample of RSGs, \cite{kiss_variability_2006} found a period of 388 $\pm$ 30 d, which was later confirmed by \cite{chatys_periodluminosity_2019}. Further works have been carried out to identify periods in Betelgeuse, for instance, using the observations from the AAVSO database between 1995 and 2014, \cite{montarges_close_2016} found a period of 423.59 $\pm$ 39 d for observations before January 2012 and a period of 422.01 $\pm$ 88 d for observations after January 2012. Furthermore, \cite{jadlovsky_analysis_2023} found a photometric period of 417 $\pm$ 17 d, while \cite{joyce_standing_2020} found a period of 416 $\pm$ 24 d and identified another shorter period of 185 $\pm$ 13.5 d being attributed to the first overtone, everything in agreement with the historical work of \cite{stothers_pulsation_1969}.}\\

Besides radial pulsation, surface activity leading to random brightness variation has also been mentioned to explain the variability of Betelgeuse, first theorised by \cite{schwarzschild_scale_1975}. More recently, \cite{gray_mass_2008} proposed that the 400 d period was a consequence of convective activity. Betelgeuse is known to present large convective cells with lifetimes of the order of one to two years \citep{lopez_ariste_convective_2018} and measurable changes 
within the span of one week. \textbf{Those typical convective timescales are also seen in numerical simulations of RSGs, where the smallest convective granules live on a timescale of a few weeks/month and the largest on a timescale of a few months up to a few years \citep{chiavassa_probing_2022}.} Bright convection cells near the disk center will increase the integrated brightness of the star compared with other situations where such cells are found near the edges. So, even without invoking the formation of 
dust or other changes in opacity, one may argue that the simple evolution of convective patterns of the star may create variability. Such 
variability could be expected to be random, but will present quasi-periodicities related to the typical time scales of the 
convective patterns \citep{gray_mass_2008}. \\

At the end of 2019, Betelgeuse reached a historical minimum in its luminosity, called the Great Dimming \citep{guinan_fall_2020}. 
From interferometric and spectroscopic data, it has been proposed that this event was caused by the formation of a cloud of dust close to the line of sight \citep{montarges_dusty_2021}. 
Interferometric images show a drop in luminosity in the southern hemisphere of Betelgeuse, which could be caused by a mass loss event and leading to this dimming \citep{dupree_great_2022}. Other hypotheses have also been explored, such as a drop in temperature \citep{harper_photospheric_2020}, or an increase in the molecular opacity \citep{kravchenko_atmosphere_2021}. In this event, it turns out that a change in brightness was not due to pulsation but to a change in the 
brightness distribution over the stellar disk.
Since the end of the Great Dimming event, Betelgeuse has continued its random variation of brightness. 
But quite interestingly, 
it has been shown by \cite{jadlovsky_analysis_2023} and \cite{dupree_great_2022} that the periodicity of Betelgeuse has changed since the dimming. Using the light curves from AAVSO, 
the cited authors showed that before the dimming, the dominant period of Betelgeuse was the 400 d period, while after the great dimming, the main periods of Betelgeuse have shortened, oscillating between 97 d and 230 d \citep{dupree_great_2022}, revealing a change in the behaviour of the atmosphere of Betelgeuse.
While the great dimming may be seen as 
a singular event, such modifications of the variability periods bring up the question of whether all changes in brightness can be due to similar changes in the brightness 
distribution over the disk and unrelated to any pulsation phenomenon. 
Our interest in an alternative explanation of the variability
in terms  of convective patterns arose. In order to address this question, we seek such typical periods of 400 and 200 d in  
observational proxies related to the convective
activity but not to any pulsation, such as the linear polarization
spectra.\\

@@

Linear polarization observed in the atomic lines is therefore a proxy of convection, a priori unrelated to radial 
pulsations but linked to the brightness inhomogeneities due to the convective patterns in Betelgeuse. If one can find the aforementioned variability periods 
in linear polarization, we can conclude that these periods are related to the convective activity which originates the linear polarization signals.
This is the purpose of the present work.\

In section \ref{Section 2}, we describe the dataset of linear polarization obtained with Narval and Neo-Narval at the 2-meter Télescope Bernard Lyot (TBL), as well as the Lomb-Scargle periodograms used to derive periods. In section \ref{section 3}, we seek periods in the \textbf{Least-Squares Deconvolution (LSD)} profiles of Betelgeuse using the \textbf{Lomb-Scargle} periodogram. In section \ref{section 4}, we associate the 200 d period with the convective timescale of the smallest granules and the 330 d periodicity with the largest granules. We speculate on an explanation in the change of variability of Betelgeuse before and after the great dimming.

@@

Betelgeuse has been observed \textbf{since 2013} with Narval and Neo-Narval at the \textbf{Télescope} Bernard Lyot\footnote[1]{https://tbl.omp.eu/}. \textbf{ The polarimeter Narval observed Betelgeuse on average every month from November 2013 to April 2019, except during summer, representing a total of 56 observations over 5.5 years. During this period, Betelgeuse was observed 43\% of the time between January and April and 57\% between September and December. Before the great dimming of Betelgeuse, Narval was replaced by Neo-Narval which is now observing Betelgeuse since 2020. } These two instruments measure the polarization 
over the visible and near infrared spectra of Betelgeuese (390-1000 nm) with high spectral resolution (R=65000) and high polarimetric sensitivity. The signal-to-noise ratios are not sufficient to measure the weak polarization signals in individual atomic lines of the 
spectrum. These amplitudes are known to be of the order of $10^{-4}$ times the continuum intensity, and they only exceptionally reach amplitudes of $10^{-3}$. When these high amplitudes are present, enough photons can be accumulated per spectral bin to 
detect the linear polarization signal above noise in lines \citep{auriere_discovery_2016}. Most commonly, the amplitudes of linear polarization are below noise levels. On those 
occasions, we have to add up the signals of thousands of lines to increase the signal-to-noise ratios. This addition 
is performed through the Least-Squares Deconvolution \citep[LSD;][]{donati_spectropolarimetric_1997,kochukhov_least-squares_2010}.
This technique has been successfully used in the past to measure 
magnetic field distributions over stellar surfaces \citep[e.g][]{donatiLandstreet2009,reiners2012,kochukhov2021} and is now employed to produce images of the brightness distributions in the photosphere 
of Betelgeuse and other RSGs, as cited in section~\ref{sec: Introduction}. \textbf{ To produce the LSD profiles, we used a solar abundance line mask, similar to the one used by \cite{auriere_discovery_2016} and \cite{mathias_evolution_2018}, which was calculated from the data provided by VALD \citep{kupka_vald2_1999}, for an effective temperature of 3750 K, $\log g =0.0$ and a microturbulence of 4 \kms, aligning with the physical paramters of \cite{josselin_atmospheric_2007}. The mask includes approximately 15,000 lines with depths exceeding 40\% of the continuum.}

@@

@@

Attempting to identify periods in an astrophysical context faces the challenge of unevenly spaced observed data. To overcome this hurdle, 
we used the Lomb-Scargle periodogram \citep{lomb_least-squares_1976,scargle_studies_1982}. This method involves fitting sinusoids to the data using least-squares with frequency sampled between the first and last observation \textbf{\citep{the_astropy_collaboration_astropy_2018}}. The quality of the fit determines the power attributed to a frequency. We applied this technique to the  LSD profiles of Stokes $I$, $U$, $Q$ and to the total linear polarization of Betelgeuse observed by the TBL over a span of 10 years (\textbf{2013}-2024). However, the dataset available post-dimming is insufficient to produce meaningful periodograms. Therefore, our focus was on periodograns computed exclusively from data obtained before the great dimming, covering the period from \textbf{ November 2013} to \textbf{April} 2019. \textbf{We only used data obtained from Narval, with a total of 56 observations}.

@@

We first computed the Lomb-Scargle periodogram at each wavelength of the LSD profile, ranging from -25~\kms to +60~\kms with a step of 1.8~\kms to cover the entire 
signal present in the LSD profile. \cite{lopez_ariste_convective_2018} identified the most blueshifted signal towards -20~\kms and interpreted it as  the maximum velocity of the rising plasma. The most redshifted signal was often found at +40~\kms, interpreted 
as the rest velocity of the star. Sometimes, signals at velocities greater than +40~\kms can be found, \textbf{with peak in the linear polarization signal up to 50~\kms \citep{lopez_ariste_three-dimensional_2022}}, which can be interpreted either as dark and cold plasma falling back to Betelgeuse or as rising plumes of plasma located on the hidden face of Betelgeuse, ascending so high that they appear above the limb \citep[for the case of the RSG $\mu$~Cep]{lopez_ariste_height_2023}. \textbf{The +60~\kms cut off in the Lomb-Scargle periodograms was set to include the occasional linear polarization peaks beyond the velocity of the star.}

Figures \ref{LS intensity},\ref{LS Q},\ref{LS U} and \ref{LS linear polarization} show the Lomb-Scargle periodogram of the LSD profile for Stokes $I$, Stokes $Q$, Stokes $U$ and the total linear polarization, respectively. 
%The abscissa is the frequency (in $\mathrm{days^{-1}}$), while the ordinate is the heliocentric radial velocity in km/s. 
In each figure, the upper panel depicts the Lomb-Scargle periodogram of each velocity bin (\textbf{gray lines}) along with the total average (red line). The green line represents the window function. \textbf{The spectral window reﬂects the pattern caused by the structure of gaps in the time string. The peaks pointed out by this window cannot be interpreted as ones related to the star.} The lower panel shows the Lomb-Scargle periodogram for each velocity bin, with the white dotted lines marking the 400 and 200 d periods for reference. The right panel represents the mean profile of each Stokes parameter at each velocity bin.

\begin{figure}[!h]
    \centering
    \includegraphics[width=0.5\textwidth]{Lomb-Scargle Intensity.pdf}
    \caption{Lomb-Scargle periodogram of the LSD profile of intensity.
    The upper panel is the Lomb-Scargle periodogram for each velocity bin (grey lines) and the average (red line). The green line is the window function.
    In the lower panel, the two white dashed lines mark respectively the 400 d and the 200 d periods. \textbf{The y-axis represents the heliocentric radial velocity (HRV). The color axis marks the power of the fit, red shows a higher power than dark blue.} The right panel is the average intensity profile for each velocity bin.}
    \label{LS intensity}
\end{figure}

@@

Examining Fig.\ref{LS intensity}, which shows the Lomb-Scargle periodogram of the LSD of the intensity profile, we observe that the 400 and 200 d periods seem to be captured by the periodogram. However, the 400 d period is uncomfortably close to the peak of the window function. Furthermore, both periods are located at the same HRV and in the blue wing of the profile. Interestingly, \cite{mathias_evolution_2018} previously found this 200 d periodicity in spectroscopic observation, despite of a shorter observation period.

Figures \ref{LS Q} and \ref{LS U} display the Lomb-Scargle periodograms of Stokes $Q$ and $U$. These periodograms exhibit significant differences compared to the intensity one. In Fig. \ref{LS Q}, a prominent signal is evident around 200 d. Regarding Stokes $U$ in Fig. \ref{LS U}, it appears that the primary \textbf{frequency} is approximately $0.003 \ \mathrm{d^{-1}}$ (equivalent to 330 d \textbf{period}). Other periods, such as those at 200 d or 250 d ($0.004 \ \mathrm{d^{-1}}$) are present, but are difficult to trust. From both periodograms of Stokes $Q$ and $U$, we recover the 200 d period, and also the 330 d period is notable, which aligns closely with the 400 d period reported in the literature and is in agreement with the timescale of the hysteresis loop reported by \cite{kravchenko_tomography_2019}.  
Figure \ref{LS linear polarization} shows the Lomb-Scargle periodogram of $\sqrt{Q^2+U^2}$, representing the total linear polarization of Betelgeuse. 
This periodogram confirms the significant powers at both 330 d and 200 d, consistent with the periodograms of Stokes $Q$ and $U$.

@@

Betelgeuse is observed, on average, every month by the TBL, enabling the tracking of its surface activity through this technique of polarimetric imaging. This technique has previously allowed for the estimation of  the size and the
 lifetime of convective cells on the surface \citep{lopez_ariste_convective_2018}. From these images, we computed a photo-center, a quantity sensitive to the size and number of convective cells. \textbf{The photo-center was computed from the equations given by \cite{chiavassa_probing_2022}.} 
A homogeneous star will have a photo-center displacement coinciding with the barycenter of the star, whereas a star with one or two large convective cells will exhibit a more siginificant photo-center displacement, up to a few percent of the stellar radius in the case of RSGs \citep{chiavassa_probing_2022}.

Since the photo-center is linked to surface convection, it is worth checking for periods in its 
dynamics over the 5 years of observations of Betelgeuse before the dimming. While these periods may overlap with those presented in the previous section, they are likely to capture additional aspects of the phenomena at work.
However, before proceeding to search for periods in the displacement of the photo-center, it is important to address a key issue regarding the interpretation of such images. 
Linear polarization suffers from a $180 ^\circ$ degree ambiguity, as mentioned in \cite{auriere_discovery_2016}. Consequently, our images 
can be 
rotated by $180^\circ$ degree, and the brightness distribution will still fit the observed LSD polarization profiles. 
If each observation were treated independently, the algorithm's solution could be any of the possible ambiguous solutions. To ensure continuity 
between the image series, 
for a given day, we use the brightness distribution of the previous day as the initial point 
of the  fitting iteration. The first image in the series begins its minimisation iteration with a random brightness distribution. 
Although we can produce a series of consistent images, it is important to keep in mind that the photo-center displacement computed from the series 
will be affected by the initial image. 
To overcome this issue, we computed
the photo-center displacement from 100 different time series, each starting from a different initial image. This approach aims to recover ensemble properties independent of the choice of the first image.

\begin{figure}[!h]
    \centering
    \includegraphics[width=0.5\textwidth]{Lomb-Scargle Photo-center.pdf}
    \caption{Lomb-Scargle periodogram of the 100 photo-center displacement series of Betelgeuse. Each blue lines correspond to a Lomb-Scargle periodogram of one photo-center
     displacement. The black line is the average of the 100 periodogram. The red dashed lines represent the 400 and 200 d period respectively.
     The green line is the window function. The black horizontal line marks the \textbf{1}$\sigma$ confidence level.}
    \label{LS photocenter}
\end{figure}

Figure \ref{LS photocenter} shows the Lomb-Scargle periodogram of the photo-center displacement for each of the 100 series (blue lines) and the average Lomb-Scargle periodogram (black line). The red dashed lines indicate the 400 and 
200 d periods, respectively. The black horizontal line represents the \textbf{1}$\sigma$ confidence level\textbf{, which was computed from the standard deviation of the average Lomb-Scarlge periodogram}. Similar to previous figures, the window function is represented by the green line. Interestingly, we find two peaks around the 400 d period in the periodogram, one at 0.0024 $\mathrm{d^{-1}}$ and one at 0.003 $\mathrm{d^{-1}}$ which are above the \textbf{1}$\sigma$ confidence level. Although they are not individually significant to confirm the presence of such period, these peaks are consistent with the findings of the previous section. Since polarimetry imaging involves only surface convection, this provides further support for a variability being explained by surface convection alone.

@@

In Fig.\ref{Lomb Scargle AAVSO}, \textbf{the upper panel is the periodogram of the mean intensity profile (red line) shown in Fig.\ref{LS intensity}. In the middle panel, we computed the periodogram from the light curve of Betelgeuse from 1990 to 2024, represented with the blue line. The orange line in lower panel depicts the periodogram computed exclusively with AAVSO observations made less than 2 days from our observation dates of the TBL. In every panels, the green line represents the window function.} For the periodogram since 1990, we binned the observations with an interval of 10 days, as Betelgeuse has been more observed in the 21st century \citep{kiss_variability_2006}. This periodogram  recovers the 400 d period and also a period close to 200 d. Other visible peaks are rather 
due to the windowing effect. \textbf{When focusing on the periodogram derived from the light curve corresponding to the TBL observation dates (lower panel), we recover the 400 d period, which is once again, very close to a peak of the window function. A smaller peak is also present around 200 d. The comparison between this periodogram and the one obtained from Stokes $I$ is important to validate our results. Stokes $I$ is linked to the total amount of light, that is the light curve. Hence, by taking observations of the AAVSO that correspond to the TBL observation dates of Betelgeuse, we should recover the same periods that the ones found in the Stokes $I$ periodogram. Since we found the 400 d and the 200 d periods in both periodograms, this validates our results. In both cases, the peak at 400 d is too close to a peak of the window function to be interpreted as a real peak. On the other hand, the second peak around 200 d, could be interpreted as a real peak. In the future, extended spectropolarimetric observations of Betelgeuse will be necessary to smooth the window function of the TBL and perhaps, disentangle the 400 d period from the window function.}

\begin{figure}[!h]
    \centering
    \includegraphics[width=0.5\textwidth]{Lomb-Scargle AAVSO.pdf}
    \caption{\textbf{Upper panel: Lomb-Scargle periodogram of the mean intensity profile presented in Fig.\ref{LS intensity} (red line).
    Middle panel: Periodogram of the light curve since 1990 as retrived from the AAVSO database (blue line).
    Lower panel: the orange line represents the periodogram of the light curve from the AAVSO when the dates correspond to the TBL observations. In every panels, the green line marks the associated window function. }}
    \label{Lomb Scargle AAVSO}
\end{figure}

\section{Discussion}
\label{section 4}

\textbf{The presence of the same 
periods in the peridogram derived from the AAVSO and the ones obtained from linear polarization and, considering the time span and sparsity of the TBL data series point to a common origin, one that is rather based upon convective dynamics.} Since we associate Stokes $Q$ and $U$ to proxies of convection, we associate the 200 d period to the typical timescales on which the Stokes parameters evolve. 
The LSD profiles of Stokes $Q$ or $U$ undergo slight changes when observing Betelgeuse monthly, but are completely overhauled 
on time scales of several months. The Lomb-Scargle periodograms seem to capture this spectral dynamic at periods of aproximately 200 d. Such changes in the profiles are directly linked to the dynamics of the surface,  to  the evolution, and movements 
of convective cells across the visible hemisphere. Therefore, the 200 d period appears to be related to the evolution of the convective patterns 
on the surface of Betelgeuse.  The stronger presence of this peak in Stokes $Q$ compared to $U$ was already pointed out by \cite{mathias_evolution_2018}, and could be interpreted as a temporal situation of how convective patterns have been distributed in the
last years.\
